---
title: "Import and Tidy NSB Supplied Location and Haul-out Data"
description: | 
  This article describes the import and tidying of location and haul-out data provided to
  MML by North Slope Burough for use in manuscript analyzing haul-out behavior of bearded, ribbon, and
  spotted seals in the Bering and Chukchi seas
date: June 14, 2021
author:
  - name: "Josh M. London"
    url: https://github.com/jmlondon
    affiliation: Alaska Fisheries Science Center
    affiliation_url: https://www.fisheries.noaa.gov/about/alaska-fisheries-science-center
    orcid_id: 0000-0002-3647-5046
output: distill::distill_article
---

```{r packages, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(tidyverse)
```

## Introduction

The focus of this document is not to analyze the data, but instead to document
the processing of the 'raw' data (as provided) into a usable data structure for
downstream merging with other datasets and analysis procedures. We will focus
on reading from the database, tidying any data structures or values that need it,
and then merging into a single usable tibble.

## Read from Database

```{r}
library(DBI)
library(odbc)
library(pingr)
library(dbplyr)
library(sf)
stopifnot(
"PEP Postgres Database Not Available; did you start VPN? ;)" = 
pingr::is_up("161.55.120.122", "5432")
)

con <- dbConnect(
  odbc::odbc(),
  dsn = "PostgreSQL pep",
  uid = keyringr::get_kc_account("pgpep_londonj"),
  pwd = keyringr::decrypt_kc_pw("pgpep_londonj")
)

deploy_qry <- "select speno, deployid, common_name as species, age_class as age, sex, tag_family, deploy_dt, end_dt
from telem.nsb_capture c
left join telem.nsb_deploy d
on c.id = d.capture_id
left join telem.lku_tag_family
using (tag_family_lku)
left join capture.lku_species
using (species_lku)
left join capture.lku_age_class
using (age_class_lku)
left join capture.lku_sex
using (sex_lku)"

nsb_deployments <- tbl(con, sql(deploy_qry)) %>% 
    dplyr::select(speno, deployid, species, age, sex, tag_family, deploy_dt, end_dt) %>% 
  collect()

nsb_deployments2 <- tibble::tribble(
  ~speno, ~deployid, ~species, ~age, ~sex, ~tag_family, ~deploy_dt, ~end_dt,
  "EB2019BW02", 149452, "Bearded seal", "Pup", "Female", "SPLASH", "2019-09-19 18:30:00", "2020-05-24 17:30:00",
  "EB2019BW02", 171333, "Bearded seal", "Pup", "Female", "SPLASH", "2019-09-19 18:30:00", "2020-08-08 06:30:00"
) %>% 
  dplyr::mutate(deploy_dt = lubridate::ymd_hms(deploy_dt),
                end_dt = lubridate::ymd_hms(end_dt))

nsb_deployments <- dplyr::bind_rows(nsb_deployments, nsb_deployments2)

nsb_timelines <- tbl(con, in_schema("telem","nsb_histos_timeline")) %>%
    dplyr::select(deployid,timeline_start_dt, percent_dry) %>% 
  collect()

nsb_timelines2 <- wcUtils::read_histos(
  here::here('data_raw/nsb/iceSeals_NSB_Alaska_vonDuyke_updatesOnly_20211008-Histos.csv'),
  dt_fmt = "%Y-%m-%d %H:%M:%S") %>% 
  wcUtils::tidyTimelines() %>% 
  dplyr::mutate(deployid = as.integer(deployid)) %>% 
  dplyr::filter(deployid %in% c(149452, 171333)) %>% 
  dplyr::select(-hist_type) %>% as_tibble()

nsb_timelines <- dplyr::bind_rows(nsb_timelines, nsb_timelines2)
  

locs_qry <- "SELECT deployid, loc_dt as locs_dt, argos_lc as quality, loc_geom as geometry
              FROM telem.nsb_locs;"

nsb_locations <- read_sf(con, query = locs_qry) %>%
    left_join(nsb_deployments, by = 'deployid') %>%
    filter(species %in% c('Bearded seal', 'Ribbon seal', 'Spotted seal')) %>%
    filter(lubridate::month(locs_dt) %in% c(3,4,5,6,7)) %>%
    rename(ptt = deployid) %>% 
    mutate(species_code = case_when(
        species == "Bearded seal" ~ "EB",
        species == "Ribbon seal" ~ "HF",
        species == "Spotted seal" ~ "PL"
    )) %>% 
    mutate(deployid = glue::glue("{species_code}",
                            "{lubridate::year(locs_dt)}_{ptt}")) %>% 
    mutate(unique_day =
             glue::glue("{lubridate::year(locs_dt)}",
                  "{lubridate::yday(locs_dt)}",
                  .sep = "_")) %>% 
  select(-species_code) %>% 
  relocate(deployid,ptt)

nsb_timelines <- nsb_timelines %>% 
  left_join(nsb_deployments, by = 'deployid') %>% 
  filter(species %in% c('Bearded seal', 'Ribbon seal', 'Spotted seal')) %>%
  filter(lubridate::month(timeline_start_dt) %in% c(3,4,5,6,7)) %>% 
  rename(ptt = deployid) %>% 
    mutate(species_code = case_when(
        species == "Bearded seal" ~ "EB",
        species == "Ribbon seal" ~ "HF",
        species == "Spotted seal" ~ "PL"
    )) %>% 
    mutate(deployid = glue::glue("{species_code}",
                            "{lubridate::year(timeline_start_dt)}_{ptt}")) %>% 
    mutate(unique_day =
             glue::glue("{lubridate::year(timeline_start_dt)}",
                  "{lubridate::yday(timeline_start_dt)}",
                  .sep = "_")) %>% 
  select(-species_code) %>% 
  relocate(deployid,ptt) %>% 
  dplyr::arrange(deployid,timeline_start_dt)

usethis::use_data(nsb_deployments, overwrite = TRUE)
usethis::use_data(nsb_timelines, overwrite = TRUE)
usethis::use_data(nsb_locations, overwrite = TRUE)
```

